A Kochen-Specker system has at least 21 vertices\\ {\small (extended abstract)}

### 2014-05-12 main.tex
Another citation, no pagenumber of the first page
Remove stray line
spellchecker

% TODO replace 21 with 22
% TODO remove this in the final version
\newcommand{\keyword}[1]{\textbf{#1}}

@@

\maketitle

\begin{abstract}
    At the heart of the Conway's Free Will theorems and Kochen and Specker's
        argument against noncontextual hidden variable theories
    is the existence of a Kochen-Specker (KS) system:
    a set of points on the sphere
    that has no~$\{0,1\}$-coloring such that
    at most one of two orthogonal points are colored~$1$
    and of three pairwise orthogonal points exactly one
    is colored~$1$.
    In public lectures, Conway encouraged the search for small
    KS systems.
    At the time of writing, the smallest known
    KS system has 31 vectors.

@@

Quantum Mechanics only predicts the probability, given the configuration
of the field, whether the particle will deviate.
Its probabilistic prediction has been thoroughly tested.
One wonders: is there a deterministic noncontextual theory predicting the
outcome of this experiment?

@@

Suppose there is a KS system and a noncontextual deterministic theory satisfying
the SPIN Axiom.
Then we color a point of this system~$0$,
whenever this theory predicts that the particle will deviate
if the spin is measured in the direction corresponding to that
point, and~$1$ otherwise.
Given two orthogonal points of the system,
we can find a third point orthogonal to both of them.
The SPIN axiom implies exactly one of them is colored~$1$, so they
cannot both be colored~$1$.
Similarly, given three pairwise orthogonal vectors in the system,
the SPIN axiom implies exactly one of them is colored~$1$.
Hence there would be a 010-coloring of the KS system, quod non.
Therefore a deterministic noncontextual theory cannot satisfy the
SPIN Axiom.

@@

\subsection{Kochen-Specker graphs}
\label{sec:ksgraphs}
We follow \cite{aow11} and reduce the search for Kochen-Specker systems
to the search of a certain class of graphs.
First note that in a Kochen-Specker system we may replace a point with its
antipodal point.  They are both orthogonal to the same points and hence
the non-010-colorability is preserved.
Therefore, we may assume antipodal points are identified on the sphere.
That is: a Kochen-Specker system is a finite subset of the projective plane
that is not 010-colorable.
\begin{dfn}
Given a finite subset~$S$ of the projective plane
(or equivalently, of the northern hemisphere without equator).
Define its \keyword{orthogonality graph}~$G(S)$ as follows.
The vertices are the points of~$S$.
Two vertices are joined by an edge, if their corresponding points
are orthogonal.
\end{dfn}
\begin{dfn}
A graph~$G$ is called~\keyword{embeddable},
if it occurs as a subgraph of an orthogonality graph;
that is: if there is a finite subset~$S$ of the projective plane,
such that~$G \leq G(S)$.
\end{dfn}
% TODO add remark on subgraph definition?
\begin{dfn}
A graph is~\keyword{010-colorable}
if there is a~$\{0,1\}$-coloring, such that
\begin{enumerate}
\item
for each triangle there is exactly one vertex that is colored~$1$ and
\item
adjacent vertices are not both colored~$1$.
\end{enumerate}
\end{dfn}
\begin{dfn}
A \keyword{Kochen-Specker graph}
is a embeddable graph that is not 010-colorable.
\end{dfn}
It is an easy, but important, consequence of the definitions that:
\begin{fact}
    A finite subset~$S$ of the projective plane
    is a Kochen-Specker system,
    if and only if its orthogonality graph~$G(S)$
    is Kochen-Specker.
\end{fact}
To prove there is no Kochen-Specker system on~$17$ points,
it would be sufficient to enumerate all graphs on~$17$ vertices
and check these are not 010-colorable or not embeddable.
However, this is infeasible as there are
already~${\sim}10^{26}$ non-isomorphic
graphs on~$17$ points.\cite{oeisA000088}
Luckily, we can restrict ourselves to certain classes of graphs.
\begin{prop}[\cite{aow11}]
    An embeddable graph is squarefree.
    That is: it does not contain the square as a subgraph.
    \footnote{Some authors call a graph squarefree if it does not
        contain the square as induced subgraph.
        For them the complete graph on four vertices is squarefree.
        We follow~Weisstein\cite{sf-weisstein} and~Sloane\cite{sf-sloane} and
        call a graph squarefree if it does not
        contain the square as subgraph.
        For us the complete graph on four vertices is not squarefree.}
\end{prop}
\begin{proof}
    Given two non antipodal points~$a\neq b$.
    Consider the points orthogonal to~$a$.
    This is a great circle.
    The points orthogonal to~$b$ is a different great circle.
    They intersect in precisely two antipodal points.
    Hence, if~$c$ and~$d$ are both orthogonal to~$a$ and~$b$,
    then~$c$ and~$d$ are equivalent.
    Therefore, an embeddable graph cannot contain a square.
\end{proof}
The squarefreeness is a considerable restriction.  There are
only~${\sim}10^{10}$ non-isomorphic squarefree graphs on~$17$ vertices.
We can restrict ourselves to connected graphs.
% TODO OEIS A006786
\begin{prop}[\cite{aow11}]\label{prop:ks-conn}
    A minimal Kochen-Specker graph is connected.
\end{prop}
\begin{proof}
    Suppose~$G$ is a non-connected Kochen-Specker graph.
    Then one of its components is not 010-colorable.
    As a subgraph of an embeddable graph, is embeddable,
    this component is embeddable as well.
    Hence it is a smaller connected Kochen-Specker graph.
\end{proof}
The gain, however, is small.
There are only~${\sim}10^9$ non-isomorphic squarefree graphs on~$17$
vertices that are not connected.
In our computations, checking for connectedness
required more time than would be gained by reducing the number of graphs.

@@

Suppose there are points in~$S'$ and~$S$
that overlap.
Note that the north pole (and south pole) is not in~$S'$. 
Let~$S''$ be~$S'$ rotated along the north pole at some angle~$\alpha$.
There are finitely many angles such that there are overlaping points.
Thus there is an angle such that~$S \cup S''$ is an embedding of~$G'$.
\end{proof}
Unfortunately, these graphs are not small.
\begin{comp}\label{comp:bic1}
Let~$F^{3,1}_n$ denote the number of connected graphs
with minimal vertex-order~$3$
on~$n$ vertices
with a vertex with fixed color~$1$.
Then:
\begin{center}
\begin{tabular}{l|lll}
    $n$ & $\leq 13$
        & $14$
        & $15$ \\
    \hline
    $F^{3,1}_n$ & $0$
        & $4$
        & $59$
\end{tabular}
\end{center}
Using the methods of Section~\ref{sec:emb},
we have determined these graphs unembeddable.
\end{comp}
We need one final computational result.
\begin{comp}\label{comp:bic2}
There are no connected graphs
on less than~$15$ vertices
with minimal vertex-order~$2$
with a vertex with fixed color~$0$. 
\end{comp}
We are ready to prove that a minimal Kochen-Specker graph
is biconnected.
\begin{proof}[Proof of Proposition~\ref{prop:ks-biconn}]
Suppose we are given a connected but not biconnected graph~$G$.
Then we can decompose
this graph into its biconnected components.  A biconnected components
is a maximal subgraph that is biconnected.
We can consider the graph of biconnected components~$B(G)$:
two biconnected components are adjacent if there is an edge
between a vertex in one component and a vertex in the other.
Note that there can at most be one edge between 
the vertices of biconnected components.
$B(G)$~does not contain loops: if it would
then the union of the biconnected components in the graph,
would be itself biconnected.  Thus~$B(G)$ is a tree.

@@

We are grateful to prof.~McKay for discussing
the feasability of certain graph restrictions.

>>>>

\usepackage{draftwatermark}

% TODO replace 21 with 22
\newcommand{\keyword}[1]{\textbf{#1}}

@@

\maketitle
\thispagestyle{empty}

\begin{abstract}
    At the heart of the Conway's Free Will theorems and Kochen and Specker's
        argument against non-contextual hidden variable theories
    is the existence of a Kochen-Specker (KS) system:
    a set of points on the sphere
    that has no~$\{0,1\}$-coloring such that
    at most one of two orthogonal points are colored~$1$
    and of three pairwise orthogonal points exactly one
    is colored~$1$.
    In public lectures, Conway encouraged the search for small
    KS systems.
    At the time of writing, the smallest known
    KS system has 31 vectors.

@@

Quantum Mechanics only predicts the probability, given the configuration
of the field, whether the particle will deviate.
Its probabilistic prediction has been thoroughly tested.
One wonders: is there a deterministic non-contextual theory predicting the
outcome of this experiment?

@@

Suppose there is a KS system and a non-contextual deterministic theory satisfying
the SPIN Axiom.
Then we color a point of this system~$0$,
whenever this theory predicts that the particle will deviate
if the spin is measured in the direction corresponding to that
point, and~$1$ otherwise.
Given two orthogonal points of the system,
we can find a third point orthogonal to both of them.
The SPIN axiom implies exactly one of them is colored~$1$, so they
cannot both be colored~$1$.
Similarly, given three pairwise orthogonal vectors in the system,
the SPIN axiom implies exactly one of them is colored~$1$.
Hence there would be a 010-coloring of the KS system, quod non.
Therefore a deterministic non-contextual theory cannot satisfy the
SPIN Axiom.

@@

\subsection{Kochen-Specker graphs}
\label{sec:ksgraphs}
We follow \cite{aow11} and reduce the search for Kochen-Specker systems
to the search of a certain class of graphs.
First note that in a Kochen-Specker system we may replace a point with its
antipodal point.  They are both orthogonal to the same points and hence
the non-010-colorability is preserved.
Therefore, we may assume antipodal points are identified on the sphere.
That is: a Kochen-Specker system is a finite subset of the projective plane
that is not 010-colorable.
\begin{dfn}
Given a finite subset~$S$ of the projective plane
(or equivalently, of the northern hemisphere without equator).
Define its \keyword{orthogonality graph}~$G(S)$ as follows.
The vertices are the points of~$S$.
Two vertices are joined by an edge, if their corresponding points
are orthogonal.
\end{dfn}
\begin{dfn}
A graph~$G$ is called~\keyword{embeddable},
if it occurs as a subgraph of an orthogonality graph;
that is: if there is a finite subset~$S$ of the projective plane,
such that~$G \leq G(S)$.
\end{dfn}
% TODO add remark on subgraph definition?
\begin{dfn}
A graph is~\keyword{010-colorable}
if there is a~$\{0,1\}$-coloring, such that
\begin{enumerate}
\item
for each triangle there is exactly one vertex that is colored~$1$ and
\item
adjacent vertices are not both colored~$1$.
\end{enumerate}
\end{dfn}
\begin{dfn}
A \keyword{Kochen-Specker graph}
is a embeddable graph that is not 010-colorable.
\end{dfn}
It is an easy, but important, consequence of the definitions that:
\begin{fact}
    A finite subset~$S$ of the projective plane
    is a Kochen-Specker system,
    if and only if its orthogonality graph~$G(S)$
    is Kochen-Specker.
\end{fact}
To prove there is no Kochen-Specker system on~$17$ points,
it would be sufficient to enumerate all graphs on~$17$ vertices
and check these are not 010-colorable or not embeddable.
However, this is infeasible as there are
already~${\sim}10^{26}$ non-isomorphic
graphs on~$17$ points.\cite{oeisA000088}
Luckily, we can restrict ourselves to certain classes of graphs.
\begin{prop}[\cite{aow11}]
    An embeddable graph is squarefree.
    That is: it does not contain the square as a subgraph.
    \footnote{Some authors call a graph squarefree if it does not
        contain the square as induced subgraph.
        For them the complete graph on four vertices is squarefree.
        We follow~Weisstein\cite{sf-weisstein} and~Sloane\cite{sf-sloane} and
        call a graph squarefree if it does not
        contain the square as subgraph.
        For us the complete graph on four vertices is not squarefree.}
\end{prop}
\begin{proof}
    Given two non antipodal points~$a\neq b$.
    Consider the points orthogonal to~$a$.
    This is a great circle.
    The points orthogonal to~$b$ is a different great circle.
    They intersect in precisely two antipodal points.
    Hence, if~$c$ and~$d$ are both orthogonal to~$a$ and~$b$,
    then~$c$ and~$d$ are equivalent.
    Therefore, an embeddable graph cannot contain a square.
\end{proof}
The squarefreeness is a considerable restriction.  There are
only~${\sim}10^{10}$ non-isomorphic squarefree graphs on~$17$
vertices.\cite{sf-sloane}
We can restrict ourselves to connected graphs.
\begin{prop}[\cite{aow11}]\label{prop:ks-conn}
    A minimal Kochen-Specker graph is connected.
\end{prop}
\begin{proof}
    Suppose~$G$ is a non-connected Kochen-Specker graph.
    Then one of its components is not 010-colorable.
    As a subgraph of an embeddable graph, is embeddable,
    this component is embeddable as well.
    Hence it is a smaller connected Kochen-Specker graph.
\end{proof}
The gain, however, is small.
There are only~${\sim}10^9$ non-isomorphic squarefree graphs on~$17$
vertices that are not connected.
In our computations, checking for connectedness
required more time than would be gained by reducing the number of graphs.

@@

Suppose there are points in~$S'$ and~$S$
that overlap.
Note that the north pole (and south pole) is not in~$S'$. 
Let~$S''$ be~$S'$ rotated along the north pole at some angle~$\alpha$.
There are finitely many angles such that there are overlapping points.
Thus there is an angle such that~$S \cup S''$ is an embedding of~$G'$.
\end{proof}
Unfortunately, these graphs are not small.
\begin{comp}\label{comp:bic1}
Let~$F^{3,1}_n$ denote the number of connected graphs
with minimal vertex-order~$3$
on~$n$ vertices
with a vertex with fixed color~$1$.
Then:
\begin{center}
\begin{tabular}{l|lll}
    $n$ & $\leq 13$
        & $14$
        & $15$ \\
    \hline
    $F^{3,1}_n$ & $0$
        & $4$
        & $59$
\end{tabular}
\end{center}
Using the methods of Section~\ref{sec:emb},
we have determined these graphs unembeddable.
\end{comp}
We need one final computational result.
\begin{comp}\label{comp:bic2}
There are no connected graphs
on less than~$15$ vertices
with minimal vertex-order~$2$
with a vertex with fixed color~$0$. 
\end{comp}
We are ready to prove that a minimal Kochen-Specker graph
is biconnected.
\begin{proof}[Proof of Proposition~\ref{prop:ks-biconn}]
Suppose we are given a connected but not biconnected graph~$G$.
Then we can decompose
this graph into its biconnected components.  A biconnected components
is a maximal subgraph that is biconnected.
We can consider the graph of biconnected components~$B(G)$:
two biconnected components are adjacent if there is an edge
between a vertex in one component and a vertex in the other.
Note that there can at most be one edge between 
the vertices of biconnected components.
$B(G)$~does not contain loops: if it would
then the union of the biconnected components in the graph,
would be itself biconnected.  Thus~$B(G)$ is a tree.

@@

We are grateful to prof.~McKay for discussing
the feasibility of certain graph restrictions.
